\documentclass{vldb}

\usepackage{amsmath,amssymb}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}

\newcommand{\comment}[1]{}
\newcommand{\tinysection}[1]{\noindent{\bf #1.}}
\newcommand{\tuple}[1]{{\langle#1\rangle}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\note}[1]{\textcolor{blue}{#1}}

\newcommand{\compiler}{DBToaster}
\newcommand{\spe}{Jasper}
\newcommand{\driver}{JDBC}

\newcommand{\targetlang}{Java}
\newcommand{\dsl}{QueryDSL}
\newcommand{\dslurl}{\url{http://www.querydsl.com}}

%\newcommand{\targetlang}{Scala}
%\newcommand{\dsl}{ScalaQuery}
%\newcommand{\dslurl}{\url{http://http://scalaquery.org}}


\toappear{}


\begin{document}
\title{Algorithmic Trading on Order Books with Embedded Queries and Synthesized
Stream Engines}
\numberofauthors{3}
\author{
\alignauthor
Oliver Kennedy\\
\affaddr{EPFL}\\
\email{oliver.kennedy@epfl.ch}
\alignauthor
Yanif Ahmad\\
\affaddr{Johns Hopkins University}\\
\email{yanif@jhu.edu}
\alignauthor
Christoph Koch\\
\affaddr{EPFL}\\
\email{christoph.koch@epfl.ch}}
\maketitle

\begin{abstract}
Algorithmic trading is an exiting area where computer science
meets economics and finance. Algorithmic
trading has gained substantial traction in recent years and is now
responsible
for a majority of the volume traded on key financial exchanges such as
NASDAQ and the London Stock Exchange.
It is an innovation engine for IT since it creates substantial
networking, data management, and computation challenges and
the players in this space have large budgets for IT infrastructure.
High-frequency algorithmic trading companies have
become a key market for high-speed switches, fast computing hardware, and data
management software -- particularly stream processors.

In this demo, we showcase a new platform for algorithmic trading.
It is based on a newly developed stream processing engine
as well as a finanical exchange simulator, a subsystem for replaying
historical data streams of buy and sell orders,
and implementations of various trading strategies.
We also demonstrate out trading strategy design and development environment.
It allows to express the data analysis portions of strategies
declaratively, using SQL on top of representations of the current
market microstructure, such as order books.
It compiles these declarative specifications down into highly efficient
code using our DBToaster compiler for {\em agile views} \cite{KAK2011},
aggressively incremental view maintenance programs that are able to
keep complex materialized SQL aggregate views continuously fresh through
updates to the base data at very high rates.
\end{abstract}

\section{Introduction}

\input{intro}


\section{The DBToaster Platform}

\input{dbtoaster}

\section{Algorithmic Trading}

\input{trading}

\section{Demo Scenario}

\todo{Demo screenshots}

\tinysection{Audience Participation}


\bibliographystyle{abbrv}
\bibliography{ref}


\end{document}
