\documentclass{sig-alternate}

\usepackage{amsmath,amssymb}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{color}

\newcommand{\longversion}[1]{}
\newcommand{\comment}[1]{}
\newcommand{\tinysection}[1]{\vspace{0.5mm}\noindent{\bf #1.}}
\newcommand{\tuple}[1]{{\langle#1\rangle}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\note}[1]{\textcolor{blue}{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}[theorem]{Example}
\def\punto{$\hspace*{\fill}\Box$}


\begin{document}
\title{DBToaster: Agile Views in a\\Dynamic Data Management System}
\numberofauthors{3}
\author{
\alignauthor
Oliver Kennedy\thanks{Also Dept.\ of Computer Science, Cornell University.}\\
%    \affaddr{\'Ecole Polytechnique F\'ed\'erale de Lausanne} \\
    \affaddr{EPFL}
%    \affaddr{Lausanne, Switzerland}
    \email{oliver.kennedy@epfl.ch}
\alignauthor
Yanif Ahmad\\
    \affaddr{Johns Hopkins University}
%    \affaddr{Baltimore, MD}
    \email{yanif@cs.jhu.edu}
\alignauthor
Christoph Koch\\
    \affaddr{EPFL}
    % \affaddr{\'Ecole Polytechnique F\'ed\'erale de Lausanne} \\
    %\affaddr{Lausanne, Switzerland}
    \email{christoph.koch@epfl.ch}
}
\maketitle

\begin{abstract}
This paper calls for a new breed of lightweight systems --
{\em dynamic data management systems (DDMS)}\/.
In a nutshell,
a DDMS manages large {\em dynamic data structures}\/ with 
{\em agile, frequently fresh views}\/, and provides a facility for monitoring
these views and triggering application-level events.
We motivate DDMS with applications in large-scale data analytics, 
database monitoring, and high-frequency algorithmic trading. 
We compare DDMS to more traditional data management systems
architectures.
%such as DBMS and data stream processors.
%
We present the DBToaster project, which is an ongoing effort to
develop a prototype DDMS system. We describe its architecture
design, techniques for high-frequency incremental view maintenance,
storage, scaling up by parallelization, and
the various key challenges to overcome to make DDMS a reality.
\end{abstract}

\input{introduction}

\section{DDMS Architecture}
\label{sec:overview}
\input{overview}


\section{Realizing Agile Views}
\label{sec:dbtoaster}
\input{dbtoaster}


\section{Managing Storage in DBToaster}
\label{sec:storage}
\input{storage}


\section{DBToaster in the Cloud}
\label{sec:distribution}
\input{distribution}

\section{Discussion and Conclusions}
\label{sec:conclusion}

We have proposed Dynamic Data Management Systems,
arguing for the need for a class of systems optimized
for keeping SQL aggregate views highly available and fresh under
high update rates.
We have sketched some of the main research challenges in
making DDMS a reality, and have outlined key design decisions
and first results and insights that make us confident that
the vision of DDMS can be realized.

We are currently developing a DDMS, DBToaster, in a collaboration
between EPFL and Johns Hopkins, which was started while the
authors worked at Cornell.
%
So far, we have developed an initial version of a
transition compiler which implements the recursive IVM technique
sketched in Section~\ref{sec:compilation}. The feasibility of
keeping views fresh through hundreds of updates per second using
this approach in the context of algorithmic trading
was recently demonstrated using an early DBToaster prototype
\cite{ahmad-vldb:09}. An initial account of the foundations and
theory of recursive IVM was given in \cite{koch-pods:10}.
Apart from substantially improving the compiler,
we are now working on the actual DBToaster DDMS. We follow the
strategy of developing two branches, one a main-memory, moderate
state-size, single-core ultra-high view refresh rate system for
applications such as algorithmic trading and the other
a cloud-based, persistent, eventual-consistency system for
very large scale interactive data analysis. We will
merge these two branches once the individual technical
problems have been solved.


\section*{Acknowledgments}


This project was supported by the US National Science Foundation under grant
IIS-0911036. Any opinions, findings, conclusions or recommendations expressed
are those of the authors and do not necessarily reflect the views of NSF.
We thank the Cornell Database Group for their feedback throughout the
development of the DBToaster project.


%\footnotesize{
\bibliographystyle{plain}
\bibliography{ref}
%}



\end{document}
