\documentclass{vldb}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{clrscode}
%\usepackage{epsfig}
%\usepackage{epic}
%\usepackage{eepic}
%\usepackage{xspace}
\usepackage{pst-tree}

%\addtolength{\textwidth}{1in}
%\addtolength{\oddsidemargin}{-0.5in}
%\addtolength{\evensidemargin}{-0.5in}
%\addtolength{\textheight}{0.8in}
%\addtolength{\topmargin}{-0.5in}
%\leftmargini 2.9ex


\def\punto{$\hspace*{\fill}\Box$}
\newcommand{\nop}[1]{}
\newcommand{\tuple}[1]{{\langle#1\rangle}}
\def\lBrack{\lbrack\!\lbrack}
\def\rBrack{\rbrack\!\rbrack}
\newcommand{\Bracks}[1]{\lBrack#1\rBrack}


\newtheorem{theorem}{Theorem}[section]
\newtheorem{metatheorem}{Metatheorem}[section]
\newtheorem{example}[theorem]{Example}
%\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{proviso}[theorem]{Proviso}
\newtheorem{todo}[theorem]{ToDo}

\newcommand{\comment}[1]{}
\newcommand{\compiler}{DBToaster}
\newcommand{\project}{Steam}

\title{DBToaster: Compiling main-memory database query~processors}

\numberofauthors{1}

\author{
\alignauthor Yanif Ahmad and Christoph Koch \\
\affaddr{Department of Computer Science} \\
\affaddr{Cornell University, Ithaca, NY} \\
\email{\{yanif, koch\}@cs.cornell.edu}}
\date{}


\toappear{}


\begin{document}


\maketitle


\begin{abstract}
Rapidly increasing main-memory capacities presents an opportunity to rethink
database system architectures. Today's databases are heavily concerned with
efficiency across the main-memory and secondary storage boundary, and their
design principles have spilled into main-memory database architectures. Moreover,
most modern database systems emphasize flexibility by interpreting queries,
flexiblity that is not required in most deployments and comes at a high
performance penalty.

We present DBToaster, a tool for compiling SQL queries into native code,
targeting main-memory databases. We present a novel compilation framework which
generates C++ code to incrementally and continuously maintain aggregate views at
high update rates using delta processing techniques. DBToaster provides two key
advantages for main-memory query processing. By generating code that performs all
the processing required for an input tuple as a straight-line function, we expose
query plan execution paths to a C++ compiler for optimization, and avoid
overheads with interpreting operator after operator. Additionally maintaining
aggregate views via standard pointer-based data structures allows us to avoid
fully computing join results without incurring prohibitive view maintenance
costs.

Our work is motivated by applications that require the highly efficient answering
of fixed workloads of aggregation queries, such as in data stream processing,
online data warehouse loading, and in financial applications. As we show, our
techniques are several orders of magnitude faster than state-of-the-art database
and stream processing engines on such workloads. In the case of queries on limit
order book data as required for supporting algorithmic equities trading, our
approach is currently the only one that can support the realistic data rates on
contemporary hardware without resorting to very substantial computing clusters.
\end{abstract}

\input{intro}
\input{overview}
\input{compilation}
\input{extensions}
\input{experiments}
\input{relwork}
\input{conc}


\footnotesize{
\bibliographystyle{abbrv}
\bibliography{ref}
}

\end{document}
