The demonstration setup will involve our algorithmic trading platform replaying
a historical order book data stream as discussed with the stock exchange
simulator in Section~\ref{sec:trading}. The demonstration will include a large
number of predefined algos, categorized by their trading methodology and scope
of operation (i.e. the exchanges and brokers they act on), to provide some
flexibility in the demo's execution. Given replayed inputs, we will
support simple pause, and seeking on the order book stream to support ad-hoc,
in-depth analysis of a snapshot of the market microstructure and algo states.

The audience will further be able to participate in the demo primarily at the
application level through the trading workstation, while still being exposed to
the technical details of query processing within the trading platform runtime.
The audience will be able to inspect the efficacy and behavior of individual
algos based on their active set of orders, add and remove algos, and
modify
their parameters and thresholds to see the effects on trades executed. The
advantages of the programming model and embedded queries will be exhibited by
flexible live editing and compilation of algos. In addition to
application-level interaction, system performance and scalability is a key
metric for our platform, with measurements such as throughput, load and latency
all present in the trading workstation visualizer alongside order book charts.
In the case of interest among conference participants, a small trading algo
competition can be performed.

\tinysection{Comparison to VLDB'09 demo}
For full disclosure, we have presented a demo at VLDB 2009 on a very early
prototype of the DBToaster compiler. Our demo workload also contained
an algorithmic trading query. However, the algorithmic trading platform
that is the object of this demonstration is entirely new; neither the
stream engine, nor any algos, nor our algo development, testing, and execution
environment we demonstrated.

%The goal of this demonstration beyond our previous in~\cite{ahmad-vldb:09} is
%threefold. First, we provide a more comprehensive look at the \compiler\
%platform, and the suitability of the tools developed to a practical application
%in algorithmic trading. This includes two usage modes: i) embedded queries, and
%the subsequent in-process query engine as opposed to a heavyweight backend
%server, and ii) direct compilation and loading into a black-box distributed
%stream engine for coarse-grained scaling. Second, while many works on streams
%and databases have referred to algorithmic trading as a source of
%queries, we have yet to see an end-to-end demonstration of a trading platform.
%A significant portion of this demonstration will be devoted to describing the
%specifics of trading algorithms and their common practices (to the best of our
%knowledge). Third, \compiler\ has significantly advanced in terms of the
%internal representations, transformations and optimization heuristics used since
%the demonstration in~\cite{ahmad-vldb:09}. This forms the foundations of the
%technical material we will present.


