\documentclass{sig-alternate}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}

\newcommand{\comment}[1]{}
\newcommand{\compiler}{Serjeant}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\feature}[1]{\textbf{#1.}}
\newcommand{\tuple}[1]{{\langle#1\rangle}}
\newcommand{\bigsum}{bigsum}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{metatheorem}{Metatheorem}[section]
\newtheorem{example}[theorem]{Example}
%\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{proviso}[theorem]{Proviso}

\graphicspath{{../figures}{./figures}}

\begin{document}
\title{Serjeant: Compiling Databases for High-Performance View Maintenance and Update Stream Processing}
\numberofauthors{2}
\author{
\alignauthor Yanif Ahmad\\
    \affaddr{Computer Science Dept.}\\
    \affaddr{Cornell University}\\
    \email{yanif@cs.cornell.edu}
\alignauthor Christoph Koch\\
    \affaddr{Computer Science Dept.}\\
    \affaddr{Cornell University}\\
    \email{koch@cs.cornell.edu}
}
\maketitle

\begin{abstract}
We present \compiler, a novel SQL query compiler for generating database engines
for high-performance main-memory processing of streaming data. \compiler\
aggressively compiles aggregate queries to C++ code, to perform incremental view
maintenance of continuous queries posed on streaming data. While today's
state-of-the-art view maintenance considers a single delta on each input
relation, \compiler\ uses a recursive compilation technique to successively
simplify queries, exploiting combinations of input deltas to eliminate joins,
and factorize aggregates. \compiler\ automatically determines subparts of
queries to eagerly maintain as map data structures during compilation, and
generates code that both computes query results from maps, and maintains maps
given their interdependencies.  \compiler\ is capable of generating incremental
engines for a wide range of aggregate queries, including nested aggregate
queries with predicates, which to the best of our knowledge has not been
addressed in the literature.  Our work is motivated by applications that require
the highly efficient answering of fixed workloads of aggregation queries, such
as in data stream processing, online data warehouse loading, and in financial
applications. As we show, our techniques are at least one to two orders of
magnitude faster than state-of-the-art database and stream processing engines on
such workloads.
\end{abstract}

\input{intro}
\input{compilation}
\input{extensions}
\input{overview}
\input{experiments}
\input{relwork}

\section{Conclusions}

\bibliographystyle{abbrv}
\bibliography{ref}

\end{document}